Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-01-15 doc/invt.tex
add modtest.c
add corr.h to compute correlation functions

In the asymptotically regime,
where $t \gg 1/\alpha(t)$,
we can approximate $\kappa_i(t)$ as
%
\begin{equation}
  \kappa_i(t) \approx \Gamma_i \, \delta(t)
  \label{eq:kappa_delta}
\end{equation}
%
where, $\Gamma_i = \int_{-\infty}^\infty \kappa_i(t) \, dt$.
%
Then, Eq. \eqref{eq:error_mbin} is simplified as
%
\begin{align}
  \mathcal E
  &=
  \int_0^t
  \sum_{i = 2}^n
  \Gamma_i \, \dot u_i^2\bigl( q(t') \bigr) \, dt',
  \label{eq:error_mbin1}
\end{align}
%
%
\note{This is equivalent to saying that the Fourier transform
  $$
  \tilde \kappa_i(\omega) = \Gamma_i
  $$
  is a roughly constant.
  %
  For the next order correction we would have
  $$
  \tilde \kappa_i(\omega) = \Gamma_i + \Gamma^{(2)}_i \omega^2 + \dots,
  $$
  which leads to a correction to $\mathcal E$
  $$
  \Gamma^{(2)}_i
  \int_0^t \ddot u_i^2\bigl( q(t') \bigr) \, dt'.
  $$
  This correction is of order $\alpha^2$
  compared to the main term in Eq. \eqref{eq:error_mbin1},
  and hence asymptotically unimportant.

  The extremal property, Eq. \eqref{eq:optimal_mbin},
  now becomes
  %
  \begin{align}
    \sum_{i=1}^n
      \Gamma_i \, \lambda_i \,
      u_i\bigl( q(\tau) \bigr) \,
      \ddot u_i\bigl( q(\tau) \bigr) = 0.
    \label{eq:optimal_mbin1}
  \end{align}
  %
  If we interpret $u_i$ as the position,
  and $\Gamma_i \, \lambda_i$ as the mass,
  then the right-hand side give the virial.
  %
  Thus, the extremal condition demands
  the total virial to be zero.
  $$\,$$
}
%
For example,
if we assume perfect sampling,
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  p_i \, (1 - p_i) \, \delta_{ij} \, \delta(t - t').
  \label{eq:zz_perfect}
\end{equation}
%
Thus, from Eqs. \eqref{eq:eig_orthonormal_rows} and \eqref{eq:eta_def},
we get
%
\begin{equation}
  \kappa_i(t - t')
  =
  \sum_{j,k = 1}^n
  \frac{ \phi_{ij} } { p_j }
  \frac{ \phi_{ik} } { p_k }
  \langle \zeta_j(t) \, \zeta_k(t') \rangle
  %
  \approx \delta(t - t').
  \label{eq:kappa_perfect}
\end{equation}
%
where we have assumed the large $n$ limit
such that $p_i \ll 1$
in the last expression.

@@

We can show from Eq. \eqref{eq:error_mbin1}
that the single-bin updating scheme
is an optimal updating scheme.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of $\beta_i \ge 0$
%
\begin{align}
&
\left( \int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \,
                     \dot u_i^2\bigl( q(t') \bigr)
\right)
\left(
\int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \, c_i^2
\right)
\notag
\\
&
\ge
\left(
\int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \, c_i \,
                     \dot u_i\bigl( q(t') \bigr)
\right)^2
\notag
\\
&
=
\left(
  \sum_{i = 2}^n \Gamma_k \, c_i
    \left[
      1 - e^{ -\lambda_i q(t) }
    \right]
\right)^2.
\notag
\end{align}
%
\note{A quick proof of the above version
  of the Cauchy-Schwarz inequality
  is the following.
  %
  First, we have
$$
\int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \,
    \left( \dot u_i \, X - c_i \right)^2
\ge 0,
$$
  for any $X$.
  %
  Thus, the right-hand side,
  as a quadratic polynomial of $X$,
  must have a non-positive discriminant.
  %
  This condition gives the desired inequality.
  \\%
}
%
Note that the last expression depends on
$q(t')$ only through the end point value,
which is fixed in the variation process.
%
The equality is achieved
if $\dot u_i\left( q(t') \right) = c_i$
for all $i = 2, \dots, n$ at any $t'$,
up to a multiplicative factor.
%
Solving this set of equations yields
$$
\alpha(t') = \frac{              1              }
                  { \lambda_i \, (t' + t_{i0} ) },
\quad
\mathrm{with\;\;}
t_{i0} = \frac{             t            }
              { e^{ \lambda_i q(t) } - 1 }.
$$
Such a solution is possible only if
\begin{equation}
\lambda_2 = \dots = \lambda_n.
\end{equation}
%
\note{
  We add some steps for solution.
  Integrating $\dot u_i = c_i$ yields
  \begin{equation}
  u_i(t') = c_i \left(t' + t_{i0} \right).
  \label{eq:ui_solution}
  \end{equation}
  Given that $u_i(t) = 1$ and $u_i(0) = e^{-\lambda_i \, q(t)}$,
  we have
  $$
  c_i = \frac{ 1 }{ t + t_{i0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{i0} } { t + t_{i0} }
  =
  e^{ -\lambda_k \, q(t) }.
  $$
  Taking the logarithm of Eq. \eqref{eq:ui_solution} yields
  $$
  -\lambda_i \, q(t) + \lambda_i \, q(t')
  = \log c_i + \log\left( t' + t_{i0} \right).
  $$
  Differentiating this with respect to $t'$ yields
  $$
  \lambda_i \, \alpha(t') = \frac{ 1 } { t' + t_{i0} }.
  $$
}

@@

We can also show that column-wise orthogonality.
%
For this we need a lemma, for $a = q \, \pi/n$,
with $q$ a positive integer.
$$
\begin{aligned}
\sum_{k = 1}^n \cos[(k - 1) \, a]
&=
1 + \cos a + \dots + \cos[(n - 1) \, a]
\\
&=
\frac{
      \sin\frac a 2
    + \sin \left[ \left( n - \frac 1 2 \right) a \right]
    }
    {
      2 \, \sin \frac a 2
    }
\\
&=
\frac{ 1 - (-1)^q } { 2 }
= \operatorname{odd}(q),
\end{aligned}
$$
%
where, we have used
$\sin \left[ \left( n - \frac 1 2 \right) a \right]
= \sin \left( q \, \pi - \frac a 2 \right)
= -(-)^q\sin\frac a 2.$
%
and $\operatorname{odd}(q)$
yields $1$ for odd $q$ or $0$ otherwise.
%
Adding the case of $q = 0$, where the sum yields $n$,
we get
$$
\sum_{k = 1}^n \cos\left[(k - 1) \, \frac { q \, \pi } { n }  \right]
= n \delta_{q, 0} + \operatorname{odd}(q).
$$

>>>>

In the asymptotically regime,
where $t \gg 1/\max \alpha(t)$,
we can approximate $\kappa_i(t)$ as
%
\begin{equation}
  \kappa_i(t) \approx \Gamma_i \, \delta(t)
  \label{eq:kappa_delta}
\end{equation}
%
where, $\Gamma_i = \int_{-\infty}^\infty \kappa_i(t) \, dt$.
%
Then, Eq. \eqref{eq:error_mbin} is simplified as
%
\begin{align}
  \mathcal E
  &=
  \int_0^t
  \sum_{i = 2}^n
  \Gamma_i \, \dot u_i^2\bigl( q(t') \bigr) \, dt',
  \label{eq:error_mbin1}
\end{align}
%
%
\note{This is equivalent to saying that the Fourier transform
  $$
  \tilde \kappa_i(\omega) = \Gamma_i
  $$
  is a roughly constant.
  %
  For the next order correction we would have
  $$
  \tilde \kappa_i(\omega) = \Gamma_i + \Gamma^{(2)}_i \omega^2 + \dots,
  $$
  which leads to a correction to $\mathcal E$
  $$
  \Gamma^{(2)}_i
  \int_0^t \ddot u_i^2\bigl( q(t') \bigr) \, dt'
  \propto
  \alpha^2 \, \mathcal E.
  $$
  This correction is of order $\alpha^2$
  compared to the main contribution in Eq. \eqref{eq:error_mbin1}.
  %
  Thus, we can assume Eq. \eqref{eq:kappa_delta},
  once $\alpha(t)$ has dropped below a certain level.
  \\
}
\note{
  The extremal property, Eq. \eqref{eq:optimal_mbin},
  now becomes
  %
  \begin{align}
    \sum_{i=1}^n
      \Gamma_i \, \lambda_i \,
      u_i\bigl( q(\tau) \bigr) \,
      \ddot u_i\bigl( q(\tau) \bigr) = 0.
    \label{eq:optimal_mbin1}
  \end{align}
  %
  If we interpret $u_i$ as the position,
  and $\Gamma_i \, \lambda_i$ as the mass,
  then the right-hand side give the virial.
  %
  Thus, the extremal condition demands
  the total virial to be zero.
  $$\,$$
}
%
For example,
if we assume perfect sampling,
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  p_i \, (1 - p_i) \, \delta_{ij} \, \delta(t - t').
  \label{eq:zz_perfect}
\end{equation}
%
Thus, from Eqs. \eqref{eq:eig_orthonormal_rows} and \eqref{eq:eta_def},
we get
%
\begin{equation}
  \kappa_i(t - t')
  =
  \sum_{j,k = 1}^n
  \frac{ \phi_{ij} } { p_j }
  \frac{ \phi_{ik} } { p_k }
  \langle \zeta_j(t) \, \zeta_k(t') \rangle
  %
  \approx \delta(t - t').
  \label{eq:kappa_perfect}
\end{equation}
%
where we have assumed the large $n$ limit
such that $p_i \ll 1$
in the last expression.

@@

We can show from Eq. \eqref{eq:error_mbin1}
that the single-bin updating scheme
is an optimal updating scheme.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of $\beta_i \ge 0$
%
\begin{align}
&
\left( \int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \,
                     \dot u_i^2\bigl( q(t') \bigr)
\right)
\left(
\int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \, c_i^2
\right)
\notag
\\
&
\ge
\left(
\int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \, c_i \,
                     \dot u_i\bigl( q(t') \bigr)
\right)^2
\notag
\\
&
=
\left(
  \sum_{i = 2}^n \Gamma_k \, c_i
    \left[
      1 - e^{ -\lambda_i q(t) }
    \right]
\right)^2.
\notag
\end{align}
%
\note{A short proof of the above version
  of the Cauchy-Schwarz inequality
  is the following.
  %
  First, we have
$$
\int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \,
    \left( \dot u_i \, X - c_i \right)^2
\ge 0,
$$
  for any $X$.
  %
  Thus, the right-hand side,
  as a quadratic polynomial of $X$,
  must have a non-positive discriminant.
  %
  This condition gives the desired inequality.
  \\%
}
%
Note that the last expression depends on
$q(t')$ only through the end point value,
which is fixed in the variation process.
%
The equality is achieved
if $\dot u_i\left( q(t') \right) = c_i$
for all $i = 2, \dots, n$ at any $t'$,
up to a multiplicative factor.
%
Solving this set of equations yields
$$
\alpha(t') = \frac{              1              }
                  { \lambda_i \, (t' + t_{i0} ) },
\quad
\mathrm{with\;\;}
t_{i0} = \frac{             t            }
              { e^{ \lambda_i q(t) } - 1 }.
$$
Such a solution is possible only if
\begin{equation}
\lambda_2 = \dots = \lambda_n.
\end{equation}
%
\note{
  We add some steps for solution.
  Integrating $\dot u_i = c_i$ yields
  \begin{equation}
  u_i(t') = c_i \left(t' + t_{i0} \right).
  \label{eq:ui_solution}
  \end{equation}
  Given that $u_i(t) = 1$ and $u_i(0) = e^{-\lambda_i \, q(t)}$,
  we have
  $$
  c_i = \frac{ 1 }{ t + t_{i0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{i0} } { t + t_{i0} }
  =
  e^{ -\lambda_k \, q(t) }.
  $$
  Taking the logarithm of Eq. \eqref{eq:ui_solution} yields
  $$
  -\lambda_i \, q(t) + \lambda_i \, q(t')
  = \log c_i + \log\left( t' + t_{i0} \right).
  $$
  Differentiating this with respect to $t'$ yields
  $$
  \lambda_i \, \alpha(t') = \frac{ 1 } { t' + t_{i0} }.
  $$
}

@@

We can also show that columnwise orthogonality.
%
For this we need a lemma, for $a = q \, \pi/n$,
with $q$ a positive integer.
$$
\begin{aligned}
\sum_{k = 1}^n \cos[(k - 1) \, a]
&=
1 + \cos a + \dots + \cos[(n - 1) \, a]
\\
&=
\frac{
      \sin\frac a 2
    + \sin \left[ \left( n - \frac 1 2 \right) a \right]
    }
    {
      2 \, \sin \frac a 2
    }
\\
&=
\frac{ 1 - (-1)^q } { 2 }
= \operatorname{odd}(q),
\end{aligned}
$$
%
where, we have used
$\sin \left[ \left( n - \frac 1 2 \right) a \right]
= \sin \left( q \, \pi - \frac a 2 \right)
= -(-)^q\sin\frac a 2.$
%
and $\operatorname{odd}(q)$
yields $1$ for odd $q$ or $0$ otherwise.
%
Adding the case of $q = 0$, where the sum yields $n$,
we get
$$
\sum_{k = 1}^n \cos\left[(k - 1) \, \frac { q \, \pi } { n }  \right]
= n \delta_{q, 0} + \operatorname{odd}(q).
$$
